---
title: "Lab 3"
author: "Charles Yang, Julia Hernesman"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readxl)
library(ggplot2)
library(tidyverse)
```

# Loading Data
```{r}
gss <- readxl::read_xlsx('GSS.xlsx')
str(gss)
```

# Cleaning
```{r}

gss %>% rename(
  year = `Gss year for this respondent`,
  happy = `General happiness`,
  party = `Political party affiliation`,
  race = `Race of respondent`,
  sex = `Respondents sex`,
  cdegree = `When received college degree`,
  hsdegree = `When received hs degree`,
  momdegree = `Mothers highest degree`,
  daddegree = `Fathers highest degree`,
  schoolyears = `Highest year of school completed`,
  age = `Age of respondent`,
  maried = `Marital status`,
  id = `Respondent id number`,
  ballot = `Ballot used for interview`) -> gss
str(gss)

gss$schoolyears <- as.numeric(gss$schoolyears)
gss$age <- as.numeric(gss$age)
```

# Average Years of Education by Policial Affiliation
```{r}
#1
by_party <- group_by(gss, party)
schoolyears_by_party <- by_party %>% summarize(schoolyears=mean(schoolyears, na.rm=TRUE))
schoolyears_by_party <- na.omit(schoolyears_by_party)
partyfactor <- factor(schoolyears_by_party$party, levels=c("Strong democrat", "Not str democrat", "Ind,near dem", "Independent", "Ind,near rep", "Not str republican", "Strong republican","Don't know", "No answer", "Other party"))

ggplot(schoolyears_by_party, aes(x=partyfactor, y=schoolyears, fill=partyfactor)) +
  geom_bar(stat="identity") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  xlab("Party") +
  ylab("Average Years of School") +
  ggtitle("Years of School by Party Affiliation") +
  ylim(0, 20) +
  scale_fill_manual("Legend", values=c("Strong democrat"="#0C00FF", "Not str democrat"="#756EFF", "Ind,near dem"="#A39EFF", "Independent"="#FF00FB", "Ind,near rep"="#FEABAB", "Not str republican"="#FC6B6B", "Strong republican"="#FF0000", "Don't know"="#BABABA", "No answer"="#666666", "Other party"="#000000"))
```

### Notes
* The differences between parties are marginal
* Strong democrats have a lesser average than Strong republicans
* Don't knows have the lowest average
* Other party has a notably high average too.

# Dataframe with better educated
```{r}
#2
betterEd <- gss %>% group_by(party) %>% filter(schoolyears > median(schoolyears, na.rm = TRUE))
str(betterEd)
```

```{r}
#3
dat1 <- gss[gss$happy %in% c('Pretty happy', 'Very happy', 'Not too happy'),]
dat2 <- dat1[dat1$maried %in% c('Married', 'Never married', 'Divorced'),]
ggplot(dat2, aes(x=maried , fill=happy)) + geom_bar(position='fill')

```

### Notes
* Being married is a better predictor for happiness

# Party Affiliation, Marital Status, and Happiness
```{r}
partyfactor <- factor(dat2$party, levels=c("Strong democrat", "Not str democrat", "Ind,near dem", "Independent", "Ind,near rep", "Not str republican", "Strong republican","Don't know", "No answer", "Other party"))

ggplot(dat2, aes(x=maried, fill=happy)) +
  geom_bar(position="fill") +
  facet_wrap(partyfactor) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  xlab("Marital Status") +
  ylab("Percent of total")
```

### Notes
* Party affiliation does not significantly affect the relationship in #3

# Marital Status by Education
```{r}
gss$schoolyrcat[gss$schoolyears < 12] <- "less education"
gss$schoolyrcat[gss$schoolyears >= 12] <- "more education"
dat3 <- gss[gss$maried %in% c('Married', 'Never married', 'Divorced'),]
dat4 <- dat3[dat3$schoolyrcat %in% c('less education', 'more education'),]
ggplot(dat4, aes(x=schoolyrcat , fill=maried)) +
  geom_bar(position='fill') +
  xlab("Education")
```

### Notes
* The more educated seem marginally happier

# Happiness by Education
```{r}
dat5 <- gss[gss$happy %in% c('Very happy', 'Pretty happy', 'Not too happy'),]
dat6 <- dat5[dat5$schoolyrcat %in% c('less education', 'more education'),]
ggplot(dat6, aes(x=schoolyrcat , fill=happy)) +
  geom_bar(position='fill') +
  xlab("Education")
```

### Notes
* There are less people "not too happy" that have more education. In other words, people that have more schooling tend to be happier.

# Father's Education and Happiness
```{r}
educatedParents <- gss[gss$daddegree %in% c("Lt high school","Bachelor","High school","Not applicable","Don't know","Graduate","Junior college"),]
and3Happy <- educatedParents[educatedParents$happy %in% c('Very happy', 'Pretty happy', 'Not too happy'),]

daddegreeFactor <- factor(and3Happy$daddegree, c("Not applicable","Lt high school","High school","Junior college","Bachelor","Graduate","Don't know"))
ggplot(and3Happy, aes(x=happy, fill=daddegreeFactor)) +
  geom_bar(position="fill")
```

### Notes
* Less happy people's dads have less education past Lt high school
* Very happy people's dads have slightly more high school education than Pretty happy people
* Pretty and Very happy are mostly the same
